---
title: "ST308 Coursework"
author: "Juhyeon Eom"
---

# Data

We use a individual-level survey microdata in Year 2021 & 2022
```{r}
install.packages("readr")
library(readr)

# Load data and select needed variables
raw_data <- read_csv("usa_00001.csv")
```

### Data cleaning
```{r}
# EDUC == 0: N/A or missing, EDUC == 1: GRADE 4 or Less, EDUC == 2: GRADE 5-8
hist(raw_data$AGE[raw_data$EDUC <= 2], breaks = 50, col = "blue")

# This shows that we can safely exclude EDUC == 0, 1, 2 -> mostly at age not eligible for working
```

```{r}
library(dplyr)

data <- raw_data %>%
  dplyr::select(SERIAL, STATEICP, SEX, AGE, MARST, RACE, EDUC, EDUCD, CLASSWKR, UHRSWORK, INCWAGE) %>%
  # INCWAGE: 999999 = N/A 999998 = Missing & Do not consider Salary-based income == 0
  filter(INCWAGE != 999999 & INCWAGE != 999998 & INCWAGE != 0) %>%
  # Exclude the case where INCWAGE is less than minimum wage based calculation
  filter(INCWAGE > 7.25 * UHRSWORK * 52) %>%
  filter(EDUC > 2) %>%
  filter(CLASSWKR != 0) %>%
  # Make a continuous variable Years of Education (YEDUC) using categorical classification (EDUC)
  mutate(YEDUC = case_when(
    EDUC == 3 ~ 9,
    EDUC == 4 ~ 10,
    EDUC == 5 ~ 11,
    EDUC == 6 ~ 12,
    EDUC == 7 ~ 13,
    EDUC == 8 ~ 14,
    EDUC == 9 ~ 15,
    EDUC == 10 ~ 16,
    # EDUCD == 114, 115 "Master's Degree" and "Professional Degree beyond Bachelor" (e.g. MBA), equivalent to two years of additional education
    EDUC == 11 & (EDUCD == 114 |EDUCD == 115) ~ 18,
    # EDUCD == 116 "Doctoral Degree", equivalent to four years of additional education
    EDUC == 11 & EDUCD == 116 ~ 22
  )) %>%
  mutate(cSEX = case_when(
    SEX == 1 ~ "Male",
    SEX == 2 ~ "Female"
  )) %>%
  mutate(cRACE = case_when(
    RACE == 1 ~ "White",
    RACE == 2 ~ "Black/African American",
    RACE == 3 ~ "American Indian or Alaska Native",
    RACE == 4 ~ "Chinese",
    RACE == 5 ~ "Japanese",
    RACE == 6 ~ "Other Asian or Pacific Islander",
    RACE == 7 ~ "Other race",
    RACE == 8 ~ "Two major races",
    RACE == 9 ~ "Three or more major races",
  )) %>%
  mutate(cMARST = case_when(
    MARST == 1 ~ "Married, spouse present",
    MARST == 2 ~ "Married, spouse absent",
    MARST == 3 ~ "Separated",
    MARST == 4 ~ "Divorced",
    MARST == 5 ~ "Widowed",
    MARST == 6 ~ "Never married/single",
    MARST == 9 ~ "Blank, missing",
  )) %>%
  mutate(STATE = case_when(
    STATEICP == 01 ~ "Connecticut",
    STATEICP == 02 ~ "Maine",
    STATEICP == 03 ~ "Massachusetts",
    STATEICP == 04 ~ "New Hampshire",
    STATEICP == 05 ~ "Rhode Island",
    STATEICP == 06 ~ "Vermont",
    STATEICP == 11 ~ "Delaware",
    STATEICP == 12 ~ "New Jersey",
    STATEICP == 13 ~ "New York",
    STATEICP == 14 ~ "Pennsylvania",
    STATEICP == 21 ~ "Illinois",
    STATEICP == 22 ~ "Indiana",
    STATEICP == 23 ~ "Michigan",
    STATEICP == 24 ~ "Ohio",
    STATEICP == 25 ~ "Wisconsin",
    STATEICP == 31 ~ "Iowa",
    STATEICP == 32 ~ "Kansas",
    STATEICP == 33 ~ "Minnesota",
    STATEICP == 34 ~ "Missouri",
    STATEICP == 35 ~ "Nebraska",
    STATEICP == 36 ~ "North Dakota",
    STATEICP == 37 ~ "South Dakota",
    STATEICP == 40 ~ "Virginia",
    STATEICP == 41 ~ "Alabama",
    STATEICP == 42 ~ "Arkansas",
    STATEICP == 43 ~ "Florida",
    STATEICP == 44 ~ "Georgia",
    STATEICP == 45 ~ "Louisiana",
    STATEICP == 46 ~ "Mississippi",
    STATEICP == 47 ~ "North Carolina",
    STATEICP == 48 ~ "South Carolina",
    STATEICP == 49 ~ "Texas",
    STATEICP == 51 ~ "Kentucky",
    STATEICP == 52 ~ "Maryland",
    STATEICP == 53 ~ "Oklahoma",
    STATEICP == 54 ~ "Tennessee",
    STATEICP == 56 ~ "West Virginia",
    STATEICP == 61 ~ "Arizona",
    STATEICP == 62 ~ "Colorado",
    STATEICP == 63 ~ "Idaho",
    STATEICP == 64 ~ "Montana",
    STATEICP == 65 ~ "Nevada",
    STATEICP == 66 ~ "New Mexico",
    STATEICP == 67 ~ "Utah",
    STATEICP == 68 ~ "Wyoming",
    STATEICP == 71 ~ "California",
    STATEICP == 72 ~ "Oregon",
    STATEICP == 73 ~ "Washington",
    STATEICP == 81 ~ "Alaska",
    STATEICP == 82 ~ "Hawaii",
    STATEICP == 83 ~ "Puerto Rico",
    TRUE ~ NA_character_
  ))

data <- na.omit(data)
summary(data)
```

Descriptive statistics (Overall)
```{r}
summary(data$INCWAGE)
summary(data$AGE)
summary(data$UHRSWORK)
summary(data$YEDUC)

sd(data$INCWAGE)
sd(data$AGE)
sd(data$UHRSWORK)
sd(data$YEDUC)
```

Descriptive statistics (Male)
```{r}
n <- data[data$cSEX == "Male",]
summary(data[data$cSEX == "Male",]$INCWAGE)
summary(data[data$cSEX == "Male",]$AGE)
summary(data[data$cSEX == "Male",]$UHRSWORK)
summary(data[data$cSEX == "Male",]$YEDUC)

sd(data[data$cSEX == "Male",]$INCWAGE)
sd(data[data$cSEX == "Male",]$AGE)
sd(data[data$cSEX == "Male",]$UHRSWORK)
sd(data[data$cSEX == "Male",]$YEDUC)
```

Descriptive statistics (Female)
```{r}
m <- data[data$cSEX == "Female",]
summary(data[data$cSEX == "Female",]$INCWAGE)
summary(data[data$cSEX == "Female",]$AGE)
summary(data[data$cSEX == "Female",]$UHRSWORK)
summary(data[data$cSEX == "Female",]$YEDUC)

sd(data[data$cSEX == "Female",]$INCWAGE)
sd(data[data$cSEX == "Female",]$AGE)
sd(data[data$cSEX == "Female",]$UHRSWORK)
sd(data[data$cSEX == "Female",]$YEDUC)
```

### Perform Exploratory Data Analysis

General Case: The effect of years of education on salary-based income (labour income)
```{r}
library(ggplot2)

colours <- c("red", "blue", "green", "yellow", "purple", "orange", "brown", "pink", "grey")

ggplot(data, aes(x=YEDUC, y=INCWAGE, fill=factor(YEDUC))) + 
  theme(
    axis.line = element_line(),
    panel.background = element_rect(fill = "white"),
  ) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values=colours) + 
  scale_y_continuous(limits = c(0, 400000),
                     expand = c(0, 0),
                     breaks = seq(0, 400000, by = 50000))
```


Income level dependent to SEX
```{r}
ggplot(data, aes(x=SEX, y=INCWAGE, fill=factor(SEX))) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(
    limits = c(0, 200000),
    expand = c(0, 0),
    breaks = seq(0, 200000, by = 25000)
  ) +
  theme_classic() + 
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.title.x = element_blank(),  
    axis.title.y = element_text(size = 12, angle = 90), 
    axis.text.x = element_text(size = 10),  
    axis.text.y = element_text(size = 10) 
  )

# 1: Male / 2: Female
```

How does the effect of education on income varies by sex
```{r}
data$SEX <- factor(data$SEX, labels = c('MALE', 'FEMALE'))

# Create the grouped box plot
ggplot(data, aes(x = factor(YEDUC), y = INCWAGE, fill = SEX)) +
  geom_boxplot(outlier.shape = NA, position=position_dodge(width=0.5), width=0.35) +
  xlab("Years of Education") +
  ylab("Salary-based Income (US$)") +
  scale_fill_manual(values = c("red", "blue")) +
  scale_y_continuous(
    limits = c(0, 200000),
    expand = c(0, 0),
    breaks = seq(0, 200000, by = 25000)
  ) +
  theme(
    axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    legend.position = "right", 
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12, angle = 90),  
    axis.text.x = element_text(size = 10), 
    axis.text.y = element_text(size = 10)  
  )

```

STATE-varient effect of education on income
```{r}
library(scales)
ggplot(data=data, aes(x=YEDUC,y=INCWAGE)) +
  stat_smooth(method="lm", formula = y ~ poly(x, 5), fullrange=TRUE, linewidth = 0.5) +
  xlab("Years of Education") + ylab("Income") + 
  scale_y_continuous(
    breaks = c(0, 50000, 100000, 150000), 
    expand = c(0, 0),
    labels = label_number(big.mark = ",")) +
  theme_minimal() + 
  facet_wrap( ~ STATE) +
  theme(axis.title=element_text(size=8),
        axis.text=element_text(size=5),
        strip.text=element_text(size=5))

```

STATE-varient income distribution
```{r}
ggplot(data, aes(x = INCWAGE, color = STATE)) + 
  geom_density() +  
  labs(title = "Probability Density of State-Varient Salary Income",
       x = "Annual Income (US$)") +
  scale_y_continuous(
    breaks = c(0, 0.000005, 0.00001, 0.000015, 0.00002), 
    expand = c(0, 0),
    labels = label_number(big.mark = ",")) +
  scale_x_continuous(
    breaks = c(0, 200000, 400000, 600000, 800000), 
    expand = c(0, 0),
    labels = label_number(big.mark = ",")) +
  theme_minimal() + 
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        plot.title = element_text(hjust = -0.05, size = 12),
        axis.title.y = element_blank(), 
        legend.position = "none")
```

# Frequentist Linear Regression Model (OLS/Ridge/Lasso)

a) OLS

```{r}
# A pooled OLS regression of income on educational attainment 
OLS1 <- lm(INCWAGE ~ YEDUC, data)
summary(OLS1)
```

```{r}
# A pooled OLS regression of income on educational attainment with controls (race, maritial status, hours of works per week, and sex)
OLS2 <- lm(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data)
summary(OLS2)
```

```{r}
# A Separate simple OLS regression of income on educational attainment with controls by SEX
OLS_male_s <- lm(INCWAGE ~ YEDUC, data[data$cSEX == "Male",])
OLS_female_s <- lm(INCWAGE ~ YEDUC, data[data$cSEX == "Female",])

summary(OLS_male_s)
summary(OLS_female_s)
```

```{r}
# A Separate OLS regression of income on educational attainment with controls by SEX, with added controls and fixed effects
OLS_male <- lm(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data[data$cSEX == "Male",])
OLS_female <- lm(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data[data$cSEX == "Female",])

summary(OLS_male)
summary(OLS_female)
```


b) Ridge & Lasso
```{r}
library(ISLR)
library(MASS)
library(glmnet)
```

Optimise tuning parameter 'lambda' using train data (cross-validation), and estimate the parameters of interest using test data

The serial number given for individual is randomly assigned - hence a sampling by a unique serial number will give randomized train and test samples for optimising lambda using cross-validation
```{r}
# Setup
data.train <- data[data$SERIAL %% 2 == 1, ]
data.test <- data[data$SERIAL %% 2 == 0, ]

train.mat <- model.matrix(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data = data.train)
test.mat <- model.matrix(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data = data.test)

grid <-  10 ^ seq(4, -2, length = 100)
```

Ridge (Frequentist)
```{r}
y <- data.train[, "INCWAGE", drop = TRUE]
mod.ridge <-  cv.glmnet(train.mat, y, alpha = 0, lambda = grid, thresh = 1e-12) 

lambda.best <-  mod.ridge$lambda.min

predict(mod.ridge, s = lambda.best, type = "coefficients")
```

Lasso (Frequentist)
```{r}
mod.lasso <- cv.glmnet(train.mat, y, alpha = 1, lambda = grid, thresh = 1e-12) 

lambda.best <-  mod.lasso$lambda.min

predict(mod.lasso, s = lambda.best, type = "coefficients")
```


# Bayesian Linear Regression Model (MCMC)

```{r}
library(rstanarm)
```

a) Simple Linear Regression, INCWAGE ~ YEDUC
```{r}
bayes1 = stan_glm(INCWAGE ~ YEDUC, data = data)
```

```{r}
bayes1
ci95_1 <- posterior_interval(bayes1, prob = 0.95)
round(ci95_1,2)
```

b) Multivariate Linear Regression, INCWAGE ~ YEDUC + controls
```{r}
bayes2 = stan_glm(INCWAGE ~ YEDUC + AGE + UHRSWORK + cRACE + cMARST + STATE, data = data)
```

```{r}
bayes2
ci95_2 <- posterior_interval(bayes2, prob = 0.95)
round(ci95_2,2)
```

c) Separate Regression by SEX
```{r}
bayes_m = stan_glm(INCWAGE ~ YEDUC + UHRSWORK + AGE + cRACE + cMARST + STATE, data = data[data$cSEX == "Male",])

bayes_f = stan_glm(INCWAGE ~ YEDUC + UHRSWORK + AGE + cRACE + cMARST + STATE, data = data[data$cSEX == "Female",])
```

```{r}
bayes_m
ci95_m <- posterior_interval(bayes_m, prob = 0.95)
round(ci95,2)
```

```{r}
bayes_f
ci95_f <- posterior_interval(bayes_f, prob = 0.95)
round(ci95,2)
```

# Multi-level Model

```{r}
install.packages("lme4")
library(lme4)
```

For the purpose of our study, a multi-model MCMC takes too long with over a million of data (over a day with random intercept alone); hence I decided to randomly select 10000 data for truncated hierarchical model
```{r}
set.seed(11)
sample_indices <- sample(nrow(data), 10000)
multidata <- data[sample_indices,]

```

Random Intercepts Model
```{r}
multi_model1 = stan_lmer(INCWAGE ~ YEDUC + SEX + (1 | STATE), data = multidata)
```

```{r}
summary(multi_model1)
```


Random Intercepts + Slopes Model
```{r}
multi_model2 = stan_lmer(INCWAGE ~ SEX + (1 + YEDUC | STATE), data = multidata)
```


```{r}
summary(multi_model2)
```


Random Slopes Model
```{r}
multi_model3 = stan_lmer(INCWAGE ~ SEX + (YEDUC | STATE), data = multidata)
```


```{r}
summary(multi_model3)
```
